### 2018-08-06 article.tex
typo and r, s, t, u to alpha, beta, gamma, delta

\input{preamble.tex}
\input{tikzpreamble.tex}
\input{quantumcircuits.tex}

@@

\begin{example}  \label{ex:quantum-circs}
  The main example of interest is \Circ, the PROP of quantum circuits.
  Different authors, and different hardware manufacturers, prefer
  different primitive gate sets but the following is typical:
  \[
  \begin{array}{cccccccccc}
    \circZ{{}} & \qquad  & \circX{{}} & \qquad & \circH \\ 
    \vstrut{1.2em}
    Z : 1 \to 1 && X : 1 \to 1 && H : 1 \to 1\\
    \\
    \circbox{R_Z(\theta)} & \qquad & \circCPhase{\theta} & \qquad & \circCX  \\ 
    \vstrut{1.2em}
    R_Z(\theta) : 1 \to 1 &&  \CPhase\theta :2 \to 2 && \CX : 2 \to 2\\
  \end{array}
  \]
  We omit the usual matrix interpretations but these can be found in
  the standard textbook \cite{NieChu:QuantComp:2000}.  Note that in
  this example we actually have an uncountable family of generating
  gates, one for each value $\theta \in [0,2\pi)$.  Unlike the case of
  Boolean circuits all quantum gates have the same number of inputs and
  outputs.  In fact, \Circ is a \dag-PROP, but this will not concern
  us here.
\end{example}

@@

\begin{definition}  \label{def:strict-two-cat}
A \emph{strict} 2-category $\mathcal{C}$ is a category \emph{enriched over
  \textbf{Cat}}, the category of categories, \emph{viz:}
\begin{itemize}
\item A category $\mathcal{C}$, whose objects and arrows are known as
  0-cells and 1-cells respectively.
\item For every pair of objects $A$ and $B$, we have a
  category $\mathcal{C}(A, B)$, whose arrows are known as 2-cells.
\item For 2-cells $\alpha : f \Rightarrow g$ and $\beta : g
  \Rightarrow h$ in $\mathcal{C}(A,B)$ their (vertical) composite is denoted
  $\beta\bullet\alpha : f \Rightarrow h$.
\item For every 1-cell $f:A\to B$ we denote its identity 2-cell by
  $Id_f:f \Rightarrow f$.
\item For all 0-cells $a$, $b$, $c$ a bifunctor $\circ: \mathcal{C}(b, c)
  \times \mathcal{C}(a, b) \to \mathcal{C}(a, c)$ called \emph{horizontal
    composition} of 2-cells.
\end{itemize}
\REM{Need here the name for a 2-cat with all 2-cells invertible;  the
  term 2-groupoid refers to the case where both 1- and 2-cells are
  invertible.}
\end{definition}

@@

\begin{definition}\label{def:monoidal-two-cat}
A \emph{strict} monoidal 2-category $(\mathcal{C}, \otimes)$ is a strict
2-category with
\REM{Is this \emph{required}?  Doesn't it follow from the  definition
  of bifunctor?  (It does in the 1-category case, so it should do
  here!)}
\begin{itemize}
\item a 2-bifunctor $\otimes: \mathcal{C} \times \mathcal{C} \to \mathcal{C}$
  such that on 2-cells $\alpha, \beta, \gamma, \delta$
\[
(\alpha \otimes \beta) \circ (\gamma \otimes \delta) = (\alpha \circ \gamma)
  \otimes (\beta \circ \delta)
\]
\REM{This definition doesn't quite work without the additoin type
  information on the 2-cells}
\end{itemize}
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        INTERCHANGE LAW REMARK        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[Interchange law]
Since $\circ$ and $\otimes$ are bifunctorial, we have the following
interchange equations for all 0-cells $A, B \in \mathcal{C}$ and all 2-cells
  $\alpha, \beta, \gamma, \delta \in \mathcal{C}(A, B)$
\begin{align}
(\alpha \circ \beta) \bullet (\gamma \circ \delta) &= (\alpha \bullet \gamma)
  \circ (\beta \bullet \delta) \\
(\alpha \otimes \beta) \bullet (\gamma \otimes \delta) &= (\alpha \bullet
  \gamma) \otimes (\beta \bullet \delta)
\end{align}
\end{remark}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        REWRITING SYSTEM DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}
Given a free PROP of quantum circuits, we can extend it to a 2-PROP
  $\mathcal{P}$ where
\begin{itemize}
\item gates form the set of 1-cells 
\item the 2-cells are generated by a finite set of \emph{rewriting} rules such
  that
\begin{itemize}
\item 2 rewriting rules with the same types must be equal
\item each rewriting rule must be invertible, that is $\mathcal{C}(A, B)$ is a
  \emph{groupoid} for all 0-cells $A, B$
\end{itemize}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        Q-ENRICHED CATEGORY DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}[$\mathbb{Q}$-enriched category]
A $\mathbb{Q}$-enriched category $\mathcal{E}$ is

\begin{itemize}
\item A set of objects
\item For all objects $(a, b)$, $\mathcal{E}(a, b) \in \mathbb{Q}$
\end{itemize}

such that for all objects $a$
\[
e \succeq \mathcal{E}(a, a)
\]
for all objects (a, b, c)
\[
\mathcal{E}(b, c) \oplus \mathcal{E}(a, b) \succeq \mathcal{E}(a, c)
\]
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        MONOIDAL Q-ENRICHED CATEGORY DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}[Monoidal $\mathbb{Q}$-enriched category]
$(E, \otimes)$ is strictly monoidal if
\[
\mathcal{E}(a, b) \oplus \mathcal{E}(a', b') \succeq \mathcal{E}(a \otimes a',
  b \otimes b')
\]
and symmetric if
\[
e \succeq \mathcal{E}(a \otimes b, b \otimes a)
\]
\end{definition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        HOMOMORPHISM DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}
For all $(A, B) \in \mathcal{P}$ we define a homomorphism $V: \mathcal{C}(A, B)
  \to \mathbb{Q}$, such that for all gates $f$
\[
V(f) \in \mathbb{Q}
\]
For all rewriting rules $\alpha: f \Rightarrow g$
\[
V(\alpha) = V(f) \oplus V(g)^{-1}
\]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        HOMOMORPHISM PROPOSITION 2        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{proposition}
The following category $\mathcal{P}(\mathbb{Q})$ is a $\mathbb{Q}$-category
\begin{itemize}
\item A set of objects which are the gates of $\mathcal{P}$
\item For all objects $(f, g)$
\end{itemize}
\[
\mathcal{E}(f, g) =
\begin{cases}
\textrm{if } f \Rightarrow g \in \mathcal{P}, &V(f \Rightarrow g) \cr
\textrm{otherwise } &\top
\end{cases}
\]
With $\top \in \mathbb{Q}$ such that for all $x \in \mathbb{Q}$
\[
x \preceq \top
\]

\begin{proof}
For all gates $(f, g)$, if it exists a unique transformation $f \Rightarrow
  g$, then
\begin{align}
\mathcal{E}(f, g) &= V(f \Rightarrow g) \in \mathbb{Q} \cr
&= V(f) \oplus V(g)^{-1}
\end{align}
otherwise we already have $\mathcal{E}(f, g) = \top \in \mathbb{Q}$.
For all gates $f$, we know that there is a identity transformation $f
  \Rightarrow f$ and
\begin{align}
\mathcal{E}(f, f) &= V(id_f) \cr
&= V(f \Rightarrow f)\cr
&= V(f) \oplus V(f)^{-1} \cr
&= e
\end{align}
Finally, for all gates $(f, g, h)$ such that there exists $\alpha: f \Rightarrow
  g$ and $\beta: g \Rightarrow h$ in $\mathcal{P}$, then the homomorphism $V$
    give us
\begin{align}
\mathcal{E}(f, h) &= V(f \Rightarrow h) \cr
&= V(\alpha \circ \beta) \cr
&= V(\alpha) \oplus V(\beta) \cr
&= V(g \Rightarrow h) \oplus V(f \Rightarrow g) \cr
&= \mathcal{E}(g, h) \oplus \mathcal{E}(f, g)
\end{align}
If either $\alpha$ or $\beta$ doesn't exists in $\mathcal{P}$ then
\[
\mathcal{E}(b, c) \oplus \mathcal{E}(a, b) \succeq \mathcal{E}(a, c)
\]
become
\[\begin{matrix}
&\mathcal{E}(b, c) \oplus \top &\succeq \top \cr
\textrm{or} &\top \oplus \mathcal{E}(a, b) &\succeq \top \cr
\textrm{or} &\top \oplus \top &\succeq \top
\end{matrix}\]
which is still true.
\end{proof}
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        HOMOMORPHISM PROPOSITION 3        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{proposition}
$\mathcal{P}(\mathbb{Q})$ is strictly symmetric monoidal with $\otimes$, the
  2-functor of $\mathcal{P}$.

\begin{proof}
The monoidal structure is immediate using the fact that
\[
V(\alpha) \oplus V(\beta) \succeq V(\alpha \otimes \beta)
\]
that is
\[
V(\alpha: f \Rightarrow g) \oplus V(\beta: f' \Rightarrow g') = V(\alpha \otimes
  \beta: f \otimes f' \Rightarrow g \otimes g')
\]
so
\[
\mathcal{E}(f, g) \oplus \mathcal{E}(f', g') = \mathcal{E}(f \otimes f', g
  \otimes g')
\]
For the symmetry, we use the fact that $V(f \Rightarrow g) \oplus V(g
  \Rightarrow f) = e$
\begin{align}
\mathcal{E}(f \otimes g, g \otimes f) &= V(f \otimes g \Rightarrow g \otimes g)
  \cr
&= V(f \Rightarrow g \otimes g \Rightarrow f) \cr
&\preceq V(f \Rightarrow g) \oplus V(g \Rightarrow f) \cr
&= V(f) \oplus V(g)^{-1} \oplus V(g) \oplus V(f)^{-1} \cr
&= e
\end{align}
\end{proof}
\end{proposition}

>>>>

\input{preamble.tex}
%%% some diagrams we won;t need in this papaer
%\input{tikzpreamble.tex}
%\input{quantumcircuits.tex}

@@

\begin{example}  \label{ex:quantum-circs}
  TODO -- quanutm circuits
\end{example}

@@

\begin{definition}  \label{def:strict-two-cat}
A \emph{strict} 2-category \catC is a category \emph{enriched over
  \textbf{Cat}}, the category of categories, \emph{viz:}
\begin{itemize}
\item A category \catC, whose objects and arrows are known as
  0-cells and 1-cells respectively.
\item For every pair of objects $A$ and $B$, we have a
  category $\catC(A, B)$, whose arrows are known as 2-cells.
\item For 2-cells $\alpha : f \Rightarrow g$ and $\beta : g
  \Rightarrow h$ in $\catC(A,B)$ their (vertical) composite is denoted
  $\beta\bullet\alpha : f \Rightarrow h$.
\item For every 1-cell $f:A\to B$ we denote its identity 2-cell by
  $Id_f:f \Rightarrow f$.
\item For all 0-cells $a$, $b$, $c$ a bifunctor $\circ: \catC(b, c)
  \times \catC(a, b) \to \catC(a, c)$ called \emph{horizontal
    composition} of 2-cells.
\end{itemize}
\REM{Need here the name for a 2-cat with all 2-cells invertible;  the
  term 2-groupoid refers to the case where both 1- and 2-cells are
  invertible.}
\end{definition}

@@

\begin{definition}\label{def:monoidal-two-cat}
A \emph{strict} monoidal 2-category $(\catC, \otimes)$ is a strict
2-category with
\REM{Is this \emph{required}?  Doesn't it follow from the  definition
  of bifunctor?  (It does in the 1-category case, so it should do
  here!)}
\begin{itemize}
\item a 2-bifunctor $\otimes: \catC \times \catC \to \catC$ such that
  on 2-cells $\alpha, \beta, \gamma, \delta$
\[
(\alpha \otimes \beta) \circ (\gamma \otimes \delta) = (\alpha \circ \gamma)
  \otimes (\beta \circ \delta)
\]
\REM{This definition doesn't quite work without the additoin type
  information on the 2-cells}
\end{itemize}
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        INTERCHANGE LAW REMARK        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[Interchange law]
Since $\circ$ and $\otimes$ are bifunctorial, we have the following
interchange equations for all 0-cells $A, B \in \catC$ and all 2-cells $\alpha,
  \beta, \gamma, \delta \in \catC(A, B)$
\begin{align}
(\alpha \circ \beta) \bullet (\gamma \circ \delta) &= (\alpha \bullet \gamma)
  \circ (\beta \bullet \delta) \\
(\alpha \otimes \beta) \bullet (\gamma \otimes \delta) &= (\alpha \bullet
  \gamma) \otimes (\beta \bullet \delta)
\end{align}
\end{remark}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        REWRITING SYSTEM DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}
Given a free PROP of quantum circuits, we can extend it to a 2-PROP $\mathbb{P}$
  where
\begin{itemize}
\item gates form the set of 1-cells 
\item the 2-cells are generated by a finite set of \emph{rewriting} rules such
  that
\begin{itemize}
\item 2 rewriting rules with the same types must be equal
\item each rewriting rule must be invertible, that is $\catC(A, B)$ is a
  \emph{groupoid} for all 0-cells $A, B$
\end{itemize}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        Q-ENRICHED CATEGORY DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}[$\mathbb{Q}$-enriched category]
A $\mathbb{Q}$-enriched category \catE is

\begin{itemize}
\item A set of objects
\item For all objects $(a, b)$, $\catE(a, b) \in \mathbb{Q}$
\end{itemize}

such that for all objects $a$
\[
e \succeq \catE(a, a)
\]
for all objects (a, b, c)
\[
\catE(b, c) \oplus \catE(a, b) \succeq \catE(a, c)
\]
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        MONOIDAL Q-ENRICHED CATEGORY DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}[Monoidal $\mathbb{Q}$-enriched category]
$(E, \otimes)$ is strictly monoidal if
\[
\catE(a, b) \oplus \catE(a', b') \succeq \catE(a \otimes a', b \otimes b')
\]
and symmetric if
\[
e \succeq \catE(a \otimes b, b \otimes a)
\]
\end{definition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        HOMOMORPHISM DEFINITION        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{definition}
For all $(A, B) \in \mathbb{P}$ we define a homomorphism $V: \catC(A, B) \to
  \mathbb{Q}$, such that for all gates $f$
\[
V(f) \in \mathbb{Q}
\]
For all rewriting rules $\alpha: f \Rightarrow g$
\[
V(\alpha) = V(f) \oplus V(g)^{-1}
\]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        HOMOMORPHISM PROPOSITION 2        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{proposition}
The following category $\mathbb{P}(\mathbb{Q})$ is a $\mathbb{Q}$-category
\begin{itemize}
\item A set of objects which are the gates of $\mathbb{P}$
\item For all objects $(f, g)$
\end{itemize}
\[
\catE(f, g) =
\begin{cases}
{\rm if~} f \Rightarrow g \in \mathbb{P}, &V(f \Rightarrow g) \cr
{\rm otherwise~} &\top
\end{cases}
\]
With $\top \in \mathbb{Q}$ such that for all $x \in \mathbb{Q}$
\[
x \preceq \top
\]

\begin{proof}
For all gates $(f, g)$, if it exists a unique transformation $f \Rightarrow
  g$, then
\begin{align}
\catE(f, g) &= V(f \Rightarrow g) \in \mathbb{Q} \cr
&= V(f) \oplus V(g)^{-1}
\end{align}
otherwise we already have $\catE(f, g) = \top \in \mathbb{Q}$.
For all gates $f$, we know that there is a identity transformation $f
  \Rightarrow f$ and
\begin{align}
\catE(f, f) &= V(id_f) \cr
&= V(f \Rightarrow f)\cr
&= V(f) \oplus V(f)^{-1} \cr
&= e
\end{align}
Finally, for all gates $(f, g, h)$ such that there exists $r: f \Rightarrow
  g$ and $s: g \Rightarrow h$ in $\catC$, then the homomorphism $V$ give us
\begin{align}
\catE(f, h) &= V(f \Rightarrow h) \cr
&= V(\alpha \circ \beta) \cr
&= V(\align) \oplus V(\beta) \cr
&= V(g \Rightarrow h) \oplus V(f \Rightarrow g) \cr
&= \catE(g, h) \oplus \catE(f, g)
\end{align}
If either $\alpha$ or $\beta$ doesn't exists in $\mathbb{P}$ then
\[
\catE(b, c) \oplus \catE(a, b) \succeq \catE(a, c)
\]
become
\[\begin{matrix}
&\catE(b, c) \oplus \top &\succeq \top \cr
{\rm or~} &\top \oplus \catE(a, b) &\succeq \top \cr
{\rm or~} &\top \oplus \top &\succeq \top
\end{matrix}\]
which is still true.
\end{proof}
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%        HOMOMORPHISM PROPOSITION 3        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{proposition}
$\mathbb{P}(\mathbb{Q})$ is strictly symmetric monoidal with $\otimes$, the
  2-functor of $\mathbb{P}$.

\begin{proof}
The monoidal structure is immediate using the fact that
\[
V(\alpha) \oplus V(\beta) \succeq V(\alpha \otimes \beta)
\]
that is
\[
V(\alpha: f \Rightarrow g) \oplus V(\beta: f' \Rightarrow g') = V(\alpha \otimes
  \beta: f \otimes f' \Rightarrow g \otimes g')
\]
so
\[
\catE(f, g) \oplus \catE(f', g') = \catE(f \otimes f', g \otimes g')
\]
For the symmetry, we use the fact that $V(f \Rightarrow g) \oplus V(g
  \Rightarrow f) = e$
\begin{align}
\catE(f \otimes g, g \otimes f) &= V(f \otimes g \Rightarrow g \otimes g) \cr
&= V(f \Rightarrow g \otimes g \Rightarrow f) \cr
&\preceq V(f \Rightarrow g) \oplus V(g \Rightarrow f) \cr
&= V(f) \oplus V(g)^{-1} \oplus V(g) \oplus V(f)^{-1} \cr
&= e
\end{align}
\end{proof}
\end{proposition}
